---
title: "Non-linear Chaotic Dynamical Systems"
author: "Par Pishrobat"
date: "`r Sys.Date()`"
format: 
  pdf:
    documentclass: article   # PDF document class (article, report, book)
    toc: true                # Table of contents for PDF
    number-sections: true    # Number sections
    toc-depth: 2        # Levels of headers to include in the TOC
    code-fold: true     # Code folding (collapsible code blocks)
    theme: cosmo        # Apply a theme (see common themes below)
  # docx:                   # For Word output
  #  toc: true
  #  reference-doc: "template.docx" # Use a custom Word template
editor: visual
#bibliography: references.bib   # Bibliography file
#csl: apa.csl              # Citation style
output-file: code_output   # Specify output filename
---

```{r include: false}
library(synlik)
invisible(lapply(list.files("R/", pattern = "\\.R$", full.names = T), source))

```

```{r}
# Ground truth parameters (relatively periodic behavior)
true_params <- c(
    alpha = 2,
    sigma = 0.05,
    y = 1
)

rulkov_sl <- synlik(
    simulator = simulate_rulkov, # Simulation function
    summaries = compute_stats, # Summary statistics function 
    param = true_params, # Initial parameters
    extraArgs = list( # Additional simulstion arguments
        "nObs"     = 50,
        "nBurn"    = 50,
        # "initVal"  = c(1,1),
        "randInit" = TRUE
    )
)

```

```{r}
rulkov_sl@data <- simulate(rulkov_sl, nsim = 1, seed = 123)
```

```{r}
rulkov_sl@plotFun <- function(input, ...) {
    plot(drop(input), type = "l", ylab = "Pop", xlab = "Time", ...)
}
plot(rulkov_sl)
```

```{r}
tmp <- simulate(rulkov_sl, nsim = 10)
dim(tmp)
```

```{r}
rulkov_sl@extraArgs$obsData <- rulkov_sl@data
```

```{r}
tmp <- simulate(rulkov_sl, nsim = 2, stats = TRUE)
tmp
```

```{r}
checkNorm(rulkov_sl)
```

```{r}
slik(rulkov_sl, param = true_params, nsim = 1e3)
```

```{r eval=FALSE}
slice(
    object = rulkov_sl,
    ranges = list(
        "alpha" = seq(1, 10, by = 0.1),
        "sigma" = seq(0.1, 1.1, by = 0.01),
        "y" = seq(0.5, 1.5, by = 0.01)
    ),
    param = true_params,
    nsim = 1000
)
```

```{r eval=FALSE}
slice(
    object = rulkov_sl,
    ranges = list(
        "alpha" = seq(1, 10, by = 0.1),
        "sigma" = seq(0.1, 1.1, by = 0.01)
    ),
    pairs = TRUE,
    param = true_params,
    nsim = 1000,
    multicore = TRUE,
    ncores = 6
)
```

```{r}
rulkov_smcmc <- synMCMC(rulkov_sl,
    initPar = c("alpha" = 1, "sigma" = 1, "y" = 1),
    niter = 51000,
    burn = 1000,
    propCov = diag(c(rep(0.1, 2), 0)),
    nsim = 100,
    priorFun = function(input, ...) mean(dunif(input)),
    targetRate = 0.15,
    multicore = FALSE
)

```

```{r}
data(rulkov_smcmc)
tmpTrans <- rep("exp", 6)
names(tmpTrans) <- names(rulkov_smcmc@param)
plot(rulkov_smcmc@data, trans = tmpTrans)
```

### Particle Marginal Metropolis–Hastings Sampler for Fast Stochastic Rulkov Map

We adapt the algorithm to fit the Fast Stochastic Rulkov Map model, where:

-   $\theta$ represents the parameters $(\alpha, y, \sigma, \eta)$.

-   $x_t$ represents the hidden (latent) state governed by the fast subsystem.

-   $z_t$ represents the observed data (noisy measurement of $x_t$).

The algorithm proceeds as follows:

### Algorithm:

#### Step 1: Initialization

-   Set iteration $i = 0$.
-   Choose an initial estimate of parameters $\theta_0 = (\alpha_0, y_0, \sigma_0, \eta_0)$.
-   Use a particle filter to estimate the likelihood $p(z^0_{1:T} | \theta_0)$.

#### Iteration $i \geq 1$:

1.  **Propose New Parameters**:\
    Sample a new set of parameters $\theta^* = (\alpha^*, y^*, \sigma^*, \eta^*)$ from the proposal distribution $K(\theta^* | \theta_{i-1})$.

2.  **Estimate Likelihood**:\
    Using a particle filter, estimate the likelihood $\hat{p}(z^0_{1:T} | \theta^*)$.

3.  **Acceptance Step**:\
    With probability: $\min \left\{ 1, \frac{\hat{p}(z^0_{1:T} | \theta^*) p(\theta^*) K(\theta_{i-1} | \theta^*)}{\hat{p}(z^0_{1:T} | \theta_{i-1}) p(\theta_{i-1}) K(\theta^* | \theta_{i-1})} \right\}$, set $\theta_i = \theta^*$. Otherwise, set $\theta_i = \theta_{i-1}$.

### 

```{r}
# Load necessary libraries
library(pomp)

# Define the state transition model (process model)
rulkov_step <- Csnippet("
  double xi = rnorm(0, sigma);
  x = alpha / (1 + x * x) + y + xi;
")

# Define the measurement model
rulkov_rmeasure <- Csnippet("
  z = rnorm(x, eta);
")

# Define the density of the measurement process
rulkov_dmeasure <- Csnippet("
  lik = dnorm(z, x, eta, give_log);
")

# Define the parameter initialization function
rulkov_init <- Csnippet("
  x = rnorm(0, 1);  // Initialize latent state
")


```

```{r}


# Simulate data based on known parameters
set.seed(123)
true_params <- c(alpha = 4.0, logeta = log(0.1), logsigma = log(0.5))

rulkov_sim <-  pomp::pomp(
    data       = data.frame(time = 1:100, obs = NA),
    times      = "time",
    t0         = 0,
    rinit      = pmcmc_init,
    rprocess   = discrete_time(pmcmc_step, delta.t = 1),
    rmeasure   = pmcmc_rmeasure,
    dmeasure   = pmcmc_dmeasure,
    dprior     = pmcmc_dprior,
    params     = true_params,
    statenames = c("x"),
    paramnames = c("alpha", "logeta", "logsigma")
  )

# Simulate observations
rulkov_sim <- simulate(rulkov_sim, nsim = 1, userdata = list(as.data.frame = TRUE))

pf_current <- pfilter(rulkov_sim, params = params, Np = 100)
loglik_current <- logLik(pf_current)
```

```{r}
library(ggplot2)
# Plot simulated data
ggplot(as.data.frame(rulkov_sim), aes(x = time, y = z)) +
    geom_line() +
    labs(title = "Simulated Observations")

```

```{r}
# Particle filter for likelihood estimation
rulkov_pfilter <- function(params, Np = 1000) {
    pfilter(
        rulkov_sim,
        params = params,
        Np = Np
    )
}

# Define the Metropolis-Hastings transition kernel with positivity constraints
mh_transition_kernel <- function(params) {
    params_new <- params
    params_new["alpha"] <- rnorm(1, mean = params["alpha"], sd = 0.1)
    params_new["y"] <- rnorm(1, mean = params["y"], sd = 0.1)

    # Ensure sigma and eta are positive by sampling from a log-normal distribution
    params_new["sigma"] <- exp(rnorm(1, mean = log(params["sigma"]), sd = 0.05))
    params_new["eta"] <- exp(rnorm(1, mean = log(params["eta"]), sd = 0.02))

    params_new
}

# Also ensure that your initial parameter guesses are valid:
params_current <- c(alpha = 3.5, logeta = log(0.2), logsigma = log(0.6))

# Particle Marginal Metropolis-Hastings algorithm
pmcmc <- function(iterations = 1000, Np = 1000) {
    # Run particle filter for initial parameter set
    pf_current <- rulkov_pfilter(params_current, Np)
    loglik_current <- logLik(pf_current)

    # Storage for results
    param_samples <- matrix(NA, nrow = iterations, ncol = length(params_current))
    loglik_samples <- numeric(iterations)

    # MCMC iterations
    for (i in 1:iterations) {
        # Propose new parameters with constrained sigma and eta
        params_new <- mh_transition_kernel(params_current)

        # Run particle filter with proposed parameters
        pf_new <- rulkov_pfilter(params_new, Np)
        loglik_new <- logLik(pf_new)

        # Check for valid log-likelihood (no NaN or -Inf)
        if (is.nan(loglik_new) || is.infinite(loglik_new)) next

        # Compute Metropolis-Hastings acceptance ratio
        accept_ratio <- exp(loglik_new - loglik_current)

        # Accept or reject
        if (runif(1) < accept_ratio) {
            params_current <- params_new
            loglik_current <- loglik_new
        }

        # Store samples
        param_samples[i, ] <- params_current
        loglik_samples[i] <- loglik_current
    }

    list(params = param_samples, loglik = loglik_samples)
}


# Run PMCMC
set.seed(123)
result <- pmcmc(iterations = 1000, Np = 1000)
```

```{r}
# Convert result to data frame for plotting
param_df <- as.data.frame(result$params)
names(param_df) <- c("alpha", "y", "sigma", "eta")

# Plot results
ggplot(param_df, aes(x = 1:1000)) +
    geom_line(aes(y = alpha), color = "blue") +
    geom_line(aes(y = y), color = "red") +
    geom_line(aes(y = sigma), color = "green") +
    geom_line(aes(y = eta), color = "purple") +
    labs(title = "Trace plots of parameter samples", x = "Iteration", y = "Parameter Value")
```
